Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-12-01 manuscript/manuscript.Rmd
added the handaxe photo as fig1.

---
title: "Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. \\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020]. This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010]. In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture. As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019]. To identify the morphological manifestation of knappers' skill level in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017]. The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].

@@

The handaxe experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 handaxes knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. These handaxes were made for previous research projects, which similarly aimed to approximate 'Late Acheulean' handaxes explicitly comparable to the Boxgrove assemblage [@faisal2010; @stout2011a; @stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 handaxes in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given. As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through [Neolithics.com](https://neolithics.com/) into flat blanks of similar size and shape for training and assessments. The mechanical properties of these raw materials are comparable to the ones used in Boxgrove in that they are both fine-grained and highly predictable in fracturing process.

@@

```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the handaxe measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
knitr::include_graphics(here::here("figure/Fig1.png"))
```

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness. It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a thicker handaxe, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip. Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.

```{r Screeplot, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
knitr::include_graphics(here::here("figure/Fig2.png"))
```

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
df1<- df1 %>% 
  mutate_if(is.numeric, round, digits = 4) %>% 
  rename(Variables = X)
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components. PC1 (Dim.1) is postively correlated with all thickness-related variables and negatively correlated with all width-related variables and the maximum length. PC2 (Dim.2) is positively with bottom width and thickness variables as well as the miximum length and negatively correlated with width and thickness variables of the tip area.") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of handaxes from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
knitr::include_graphics(here::here("figure/Fig3.png"))
```

@@

```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132). The upper left area in each individual plot displays statistical reporting from a frequentist perspective, including the student-t test statistics, p-value, Pearson correlation coefficient, confidence interval, and sample size. The lower right area in each individual plot displays statistical reporting from a Bayesian perspective, including the natural logarithm of Bayes factor, posterior type and estimate, credible interval, and prior type and value. "}
knitr::include_graphics(here::here("figure/Fig4.png"))
```

@@

```{r PCA1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC1 values."}
knitr::include_graphics(here::here("figure/Fig5.png"))
```

```{r PCA2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC2 values."}
knitr::include_graphics(here::here("figure/Fig6.png"))
```

@@

The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)). If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005]. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)). It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape. Experts generally try to achieve handaxe forms while removing as little mass as possible (i.e. making as big a handaxe as possible from the nodule) [@whittaker1994]. On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, we infer that behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the knapping process."}
knitr::include_graphics(here::here("figure/Fig7.jpg"))
```

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.

>>>>

---
title: "Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. \\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020]. This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010]. In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)). As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019]. To identify the morphological manifestation of knappers' skill level in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017]. The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].

```{r photos, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A selection of Boxgrove handaxes and modern replicas produced by experts and novices."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
```

@@

The handaxe experimental replicas used in this study comprised two sub-collection (**Figure** \@ref(fig:photos)). The first sub-collection includes 10 handaxes knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. These handaxes were made for previous research projects, which similarly aimed to approximate 'Late Acheulean' handaxes explicitly comparable to the Boxgrove assemblage [@faisal2010; @stout2011a; @stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 handaxes in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given. As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through [Neolithics.com](https://neolithics.com/) into flat blanks of similar size and shape for training and assessments. The mechanical properties of these raw materials are comparable to the ones used in Boxgrove in that they are both fine-grained and highly predictable in fracturing process.

@@

```{r ImageJ, echo=FALSE, out.width = '80%', out.height = '50%', fig.align='center', fig.cap="A visual demonstration of the handaxe measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
knitr::include_graphics(here::here("figure/Fig2.png"))
```

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness. It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a handaxe with thicker shape relative to width and length, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip. Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.

```{r Screeplot, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
knitr::include_graphics(here::here("figure/Fig3.png"))
```

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
df1<- df1 %>% 
  mutate_if(is.numeric, round, digits = 4) %>% 
  rename(Variables = X)
df1[1, 1] <- "width_90%"
df1[2, 1] <- "width_80%"
df1[3, 1] <- "width_70%"
df1[4, 1] <- "width_60%"
df1[5, 1] <- "width_50%"
df1[6, 1] <- "width_40%"
df1[7, 1] <- "width_30%"
df1[8, 1] <- "width_20%"
df1[9, 1] <- "width_10%"
df1[10, 1] <- "thickness_90%"
df1[11, 1] <- "thickness_80%"
df1[12, 1] <- "thickness_70%"
df1[13, 1] <- "thickness_60%"
df1[14, 1] <- "thickness_50%"
df1[15, 1] <- "thickness_40%"
df1[16, 1] <- "thickness_30%"
df1[17, 1] <- "thickness_20%"
df1[18, 1] <- "thickness_10%"
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components. PC1 (Dim.1) is postively correlated with all thickness-related variables and negatively correlated with all width-related variables and the maximum length. PC2 (Dim.2) is positively with bottom width and thickness variables as well as the miximum length and negatively correlated with width and thickness variables of the tip area.") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of handaxes from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
knitr::include_graphics(here::here("figure/Fig4.png"))
```

@@

```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132). The upper left area in each individual plot displays statistical reporting from a frequentist perspective, including the student-t test statistics, p-value, Pearson correlation coefficient, confidence interval, and sample size. The lower right area in each individual plot displays statistical reporting from a Bayesian perspective, including the natural logarithm of Bayes factor, posterior type and estimate, credible interval, and prior type and value. "}
knitr::include_graphics(here::here("figure/Fig5.png"))
```

@@

```{r PCA1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC1 values."}
knitr::include_graphics(here::here("figure/Fig6.png"))
```

```{r PCA2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC2 values."}
knitr::include_graphics(here::here("figure/Fig7.png"))
```

@@

The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)). If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113]. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** @ref(fig:weight)). It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape. Achieve handaxe forms while removing as little mass as possible (i.e. making as big a handaxe as possible from the nodule) generally requires a higher skill level due to the reductive or subtractive nature of stone knapping, where correcting an error or any thinning procedure always requires the removal of raw material and thereby reducing the size of a given handaxe [@schillingerCopyingErrorCultural2014a: 130; @deetz1967: 48-49]. On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, we infer that behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the knapping process."}
knitr::include_graphics(here::here("figure/Fig9.jpg"))
```

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.

### 2022-12-01 manuscript/manuscript_word.Rmd
added the handaxe photo as fig1.

---
title: 'Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights'
author:
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY,  USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
output: word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

@@

```{r Screeplot, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
knitr::include_graphics(here::here("figure/Fig2.png"))
```

@@

```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the knapping process."}
knitr::include_graphics(here::here("figure/Fig7.jpg"))
```

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizibility to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.

>>>>

---
title: 'Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights'
author:
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY,  USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
output:
  word_document:
    reference_docx: "manuscript word.docx"
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

@@

```{r Screeplot, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
knitr::include_graphics(here::here("figure/Fig3.png"))
```

@@

```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the knapping process."}
knitr::include_graphics(here::here("figure/Fig9.jpg"))
```

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.
